Factors determining how catchments respond to forest cover change: Re-analysing global data sets.

### 2022-01-12 paper/Forest_and_Water/Forest_and_Water.Rmd
updates on paper 12012022

---
title: Do larger watersheds respond different to forest cover change? Re-analysing a global data set.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357 Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.
  It consists of two paragraphs.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017]. The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian watersheds) a 100% decrease in streamflow for watersheds with 100% forest cover. However, on the other end of the spectrum, in a series of French watersheds [@cosandey2005], there was no change in streamflow characteristics in 2 of the three watersheds studied in relation to deforestation.

@@

Several review papers have summarized different studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017]. These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of watershed studies in relation to changes in streamflow due to changes in forest cover. The @zhang2017 dataset, which covers over 250 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change.

@@

The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a relationship between watershed size and decline in streamflow.

@@

Encouraged by the work presented by @zhang2017 and @filoso2017 and the fantastic database of studies presented by these authors, we believe we can add to the discussion. In this paper, the aim is to develop further analysis of the collected data and expanding and combining the two data sets to provide further depth.

In particular, the main method in the work by @zhang2017 is using simple linear regression, and in @filoso2017 the focus is mainly on classification. As @zhang2017 points out, the main assumption in their work is that the threshold at 1000 km^2^ is a distinct separation between "small" and "large" watersheds, but the subset of data in @filoso2017 does not appear to support this. And while te work @filoso2017 provides important insights in study types, analysis types and broad classification, there is limited quantification of actual impact. This is because the work had a strict criterion to select quantitative studies. However, given the fantastic data sets collected, the analyses can be easily expanded to look at interactions between the terms and to test the assumption of a distinct threshold at 1000 km^2^.

As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further watersheds (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear, interactions and partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).

Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is clearly modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f. @jackson2005) and possibly the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).

@@

While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watershed data sets in our analysis.

@@

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for watershed #76 was corrected to 157% based on the original publication [@baker1984]. Finally, on review of all the data in @zhang2017, 29 potential duplicates were identified and flagged in the data.

@@

$$\tag{2}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}$$

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` observations. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42\% of the data from watersheds < 1 km^2^ and 65\% of the data for watersheds < 10 km^2^.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Flow (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

(ref:datagraphscaption) foo

@@

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("increase" = "blue", "decrease" = "red")) +
  xlab("Change in forest cover %") + ylab("Change in flow %")
dev.off()

@@

$$\tag{3}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}$$

@@

$$\tag{4}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
\end{aligned}$$

@@

$$\tag{5}
\begin{aligned}
\Delta \% Qf \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response. This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005). In addition to adding the length as a variable, all continuous variables (Dryness, Area, length and Latitude) were considered non-linear in this model and as indicated a shrinkage smoothing spline [@wood2006] was applied to these variables.

@@

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
                    s(length, bs="ts") ,
                  data = All_data2)
#summary(model5_all)
#gam.check(model5_all)
#plot(model5_all)
```

```{r mfive-linear, echo = F}
pander(round(summary(model5_all)$p.table,2), caption = "Statistical summary for the linear terms in the model with non-linear terms")
```

```{r mfive-smooth, echo = F}
pander(round(summary(model5_all)$s.table,2), caption = "Statistical summary for the smooth terms in the model with non-linear terms")

@@

<!-- model 2c include smooth variables -->

@@

A final full model which includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime) is tested.

@@

model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts" ) + 
                     s(length, bs="ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime
                    , data = All_data2)
#summary(model6_all)
#gam.check(model6_all)
#plot(model6_all)
```

```{r msix-linear, echo = F}
pander(round(summary(model6_all)$p.table,2), caption = "Statistical summary for the linear terms the full model")
#plot(model6_all)
```

```{r msix-smooth}
pander(round(summary(model6_all)$s.table,2), caption = "Statistical summary for the smooth terms for the full model")
#plot(model6_all)
```

@@

```{r visualisesmooths,echo =F, eval = F}
#require(visreg)
png("model6_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
D / (L + A) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```

@@

Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area) and the change in flow is essentially linear and does not need to be smoothed (this is the value of using penalized smooths following @wood2006). It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds. In addition, both the length and Dryness variables show non-linearity, but this does not show a clear trend.

@@

<!--filter lenght < 60 -->
```{r model7_noLatb, echo = F, eval = F}
model7_nolongLength <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs="ts") + s(length,bs="ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = All_data2 %>% filter(length< 60))
#summary(model7_noLatb)
#plot(model7_noLatb)
#pander(round(summary(model7_nolongLength)$p.table,2))
#plot(model7_noLatb)
#summary(model7_nolongLength)
```
```{r restrictlength, echo = F}
pander(round(summary(model7_nolongLength)$s.table,2), caption = "Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
```

The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured. The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005. One reason could be the few data points with very long data series, but highly variable responses (Figure \@ref(fig:smoothsmodel6)). Reducing the flexibility of the splines, removing any studies longer than 60 years (Table \@ref(tab:restrictlength)), or fitting a linear term results in "length" not being significant, suggesting that this variable is not helpful in explaining the variation in the data.

@@

```{r, eval =F, echo =F}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```
```{r, eval =F}
Zhang_all2 %>%
  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
```

@@

@@

Essentially, the analysis shows at the moment that in contrast to @zhang2017 there is no evidence that the size of a watershed influences the change in the streamflow as a result of changes in forestry. If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds. In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.

As shown earlier, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds. This is a confounding factor that is not easily addressed in the regression modelling attempted here. Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.

### remove the assessment techniques with very small numbers

One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis. This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM).

@@

pander(Table_assessment)
```

>>>>

---
title: Factors determining how catchments respond to forest cover change: Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE, The University of Sydney
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and the Environment
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 357, Instituto Nacional de Investigación Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Deforestation causes a 32\% greater change in flow compared to reforestation. Area of the catchment only affects the change in streamflow after log transformation, given the wide variety in the data from small scale paired watershed studies. Smaller studies dominate the database with 42\% of the data < 1 km2 and 65\% of the data < 10 km2. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38/%) is low due the quality of the inputs and additional unknown confounding factors.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

There has been an long and on-going discussion in the hydrologcal literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017]. The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchmensts after deforestation.

@@

Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017]. These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover. The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets.

@@

The second study [@filoso2017] was a systematic review which classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of quantitative data (37) they showed a log-linear relationship between decreasing watershed size and an increasing decline in streamflow. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

@@

Encouraged by the work presented by @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion. In this paper, the aim extend the analysis of the collected data and to expand and combine the data sets.

In particular, the main method in the work by @zhang2017 is a single covariate linear regression, and in @filoso2017 the focus is mainly on classification and there is again some single covariate linear regression. As @zhang2017 points out, a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments. However, the subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum. And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.

As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further catchments (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).

Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f. @jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).

@@

While @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration was not originally included as part of the published data set.
We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) catchment data sets in our analysis.

@@

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for watershed #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data.

@@

$$\tag{2}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & \sum{X_i} + \sum{s(Z_i)} + \varepsilon
\end{aligned}$$

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` watersheds and the @filoso2017 study used 37 watersheds (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42\% of the data from watersheds < 1 km^2^ and 65\% of the data for watersheds < 10 km^2^.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

(ref:datagraphscaption) Overview of the distribution of the data set for five of the included variables.

@@

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest change direction", 
                      values = c("increase" = "blue", "decrease" = "red")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

@@

$$\tag{3}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon
\end{aligned}$$

@@

$$\tag{4}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + \varepsilon
\end{aligned}$$

@@

$$\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & Pa_{mm} + log10(Area_{km^2}) + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response. This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005). In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, three other continuous variables (*Dryness, Area, From*) were considered non-linear in this model. This is also indicated a shrinkage smoothing spline [@wood2006] was applied to these variables. *From* represents the starting date of the data collection.

$$\tag{5}
\begin{aligned}
\Delta Qf \% \sim ~ &\Delta \% forest~cover_{positive} + sign_{forest~cover} + \\ & s(log10(Area ~(km^2)) + s(length) + s(Dryness) + \\ & s(From) + \varepsilon
\end{aligned}$$

<!--
```{r model4b, echo = F, eval = F}
All_data2 <- All_data2 %>%
  mutate(length = To - From)

model4b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + From + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") +
                 s(length, bs = "ts"),
                  data = All_data2 %>% filter(length < 60))
summary(model4b)
#gam.check(model5_all)
#plot(model5_all)
```
-->

@@

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts") + 
                    s(length, bs="ts") + s(From, bs = "ts") ,
                  data = All_data2)
#summary(model5_all)
#gam.check(model5_all)
#plot(model5_all)
```

```{r , echo = F}
pander(round(summary(model5_all)$p.table,2), caption = "(\\#tab:mfive-linear) Statistical summary for the linear terms in the model with non-linear terms")
```

```{r , echo = F}
pander(round(summary(model5_all)$s.table,2), caption = "(\\#tab:mfive-smooth) Statistical summary for the smooth terms in the model with non-linear terms")

@@

@@

A final model includes the remaining categorical variables (Precipitation data type, Assessment technique, Forest type and Hydrological regime).

@@

model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(log10(Area_km2), bs="ts") + s(Dryness, bs="ts" ) + 
                     s(length, bs="ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime
                    , data = All_data2)
#summary(model6_all)
#gam.check(model6_all)
#plot(model6_all)
```

```{r, echo = F}
pander(round(summary(model6_all)$p.table,2), caption = "(\\#tab:msix-linear) Statistical summary for the linear terms the full model")
#plot(model6_all)
```

```{r , echo = F}
pander(round(summary(model6_all)$s.table,2), caption = "(\\#tab:msix-smooth) Statistical summary for the smooth terms for the full model")
#plot(model6_all)
```

@@

```{r visualisesmooths,echo =F, eval = F}
#require(visreg)
png("model6_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model6_all, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model6_all, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model6_all, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model6_all, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```

@@

Figure \@ref(fig:smoothsmodel6) highlights that the relationship between log10(Area km^2^) and the change in flow is essentially linear, not significant, and does not need to be smoothed (this is the value of using penalized smooths following @wood2006). It is still a negative slope, indicating that in larger watersheds the impact of changes in forest cover on streamflow is less than for smaller watersheds. Similarly, the length variable is not significant. However, both the length and Dryness variables show non-linearity, but this does not show a clear trend due to the scatter and the distributions of the data.

@@

<!--filter lenght < 60 -->
```{r model7_nolongLength, echo = F}
NoLong_Data <- All_data2 %>% filter(length< 60, Dryness <= 4)
model7 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + #s(Latitude, bs="ts") + 
                      s(log10(Area_km2), bs = "ts") + s(length, bs = "ts") + s(From, bs = "ts") +
                    Precip_data_type +  Assessment_technique + Forest_type +
                    Hydrological_regime, data = NoLong_Data)
#summary(model7)
```
```{r, echo = F}
pander(round(summary(model7)$s.table,2), caption = "(\\#tab:restrictlength) Statistical summary of the smooth terms reducing dataset to studies with the study length shorter than 60 years")
```
```{r visualisesmooths7,echo =F, eval = F}
#require(visreg)
png("model7_smooths.png", width = 2500, height = 1800, res = 300)
#par(mfrow=c(2,2))
D <- visreg(model7, "Dryness", gg = T, plot = T) + theme_bw()
A <- visreg(model7, "Area_km2", gg = T, plot = T) + theme_bw() + scale_x_log10()
L <- visreg(model7, "length", gg = T, plot = T) + theme_bw()
From <- visreg(model7, "From", gg = T, plot = T) + theme_bw()
(D + A) / (L + From) 
# visreg(model6_all,"Dryness",gg=T) + theme_bw()
# visreg(model6_all,"length",gg=T, add=T) + theme_bw()
# visreg(model6_all,"Area_km2",gg=T, add=T) + theme_bw() + scale_x_log10()
#par(mfrow=c(1,1))
dev.off()
```
```{r smoothsmodel7, fig.cap="Visualisation of the smooth variables in the model wiht reduced data for dryness and length", out.width="90%", echo=F}

include_graphics("model7_smooths.png")
```

The flexible nature of the splines means that the length variable captures some substantial variation in the data, but it is unclear what exactly is captured. The shape of the conditional response (Figure \@ref(fig:smoothsmodel6)) also does not reflect the type of response highlighted in @filoso2017 and @jackson2005. One reason could be the few data points with very long data series, and very old studies (before 1930 essentially), and highly variable responses (Figure \@ref(fig:smoothsmodel6)). Reducing the flexibility of the splines, removing any studies longer than 60 years and removing the 4 studies with Dryness > 4 (Table \@ref(tab:restrictlength)) result in log10(Area) showing a significant effect with larger catchments showing less impact of changes in forest cover. Dryness now shows a clear non-linear response with drier catchments having a greater impact of changes in forest cover. Finally, more recent studies also show greater impacts of changes in forest cover (Figure \@ref(fig:smoothsmodel7)).

@@

```{r, eval =F, echo =F}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```
```{r, eval =F, echo = F}
Zhang_all2 %>%
  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
```

@@

```{r overview, fig.cap="Overview of the data highlighting the dominance of small catchment studies which are fully forested or cleared and the scatter in the data", out.width="90%", echo=F}

include_graphics("flow_forest_byArea.png")

@@

Essentially, the analysis shows that in contrast to @zhang2017 there is no evidence a distinct threshold in the size of the catchment that influences the change in the streamflow as a result of changes in forestry. If anything the scatter in the data (in the change in flow) is greater for the smaller watersheds then for the larger watersheds (Figure \@ref(fig:overview)). In other words, the response to changes in forest cover is more consistent for larger watersheds than it is for smaller watersheds.

However, most of the smaller watersheds are "real observed data" using paired watershed studies, while for larger watersheds, the analysis are mostly based on modelling approximations using either elasticity analysis (EA), Hydrological modelling (HM) or a combined use of statistical methods (SH) or quasi paired watershed analysis (QPW) (Figure \@ref(fig:overview)), thus all providing an approximation of the effect of forestry on streamflow rather than a direct comparison of watersheds. This is a confounding factor that is not easily addressed in the regression modelling attempted here. Furthermore, the catchments analysed using EA, are concentrated in the drier end of the Dryness index scale compared to the other methods, with only the paired watershed experiment (PWE) assessment technique covering the full range of dryness indices.

## The effect of assessment techniques with very small numbers of observations

One concern is that there are a few Assessment techniques in the original dataset with a very low number of observations and this might skew the results of the analysis. This includes the category of Quasi paired watersheds and combinations of elasticity analysis and hydrological modelling (EA,HM) and paired watersheds and hydrological modelling (PWE,HM). The following model removes these from the data set.

@@

pander(Table_assessment, caption = "(\\#tab:tableassess) Distribution of assessment techniques in the data set")
```

```{r, eval = F, echo =F}
png("AssessmentTechnique_byArea.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  ggplot(aes(Assessment_technique, log10(Area_km2))) + 
  geom_boxplot(colour = "blue") + theme_bw() +
    theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size = rel(1.5)))
dev.off()
```
